---
title: "ICA-Based Artifact Removal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ICA-Based Artifact Removal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

Independent Component Analysis (ICA) is a blind source separation technique
commonly used to identify and remove artifacts from electrophysiological
recordings. Artifacts such as eye blinks, eye movements, muscle activity, and
heartbeat contamination can be isolated as independent components and removed
without discarding entire channels or epochs.

PhysioPreprocess provides two ICA workflows:

1. **Built-in ICA** (`icaDecompose()` / `icaRemove()`): A self-contained
   FastICA implementation with no additional dependencies.
2. **fastICA-based** (`runICA()` / `removeICAComponents()`): A wrapper around
   the `fastICA` package for users who prefer that interface.

This vignette walks through both approaches.

## Setup

```{r load-packages}
library(PhysioCore)
library(PhysioPreprocess)
```

## Creating Example Data

We simulate a multi-channel EEG recording containing a mixture of neural
sources and artifact sources.

```{r create-data}
set.seed(123)
n_time <- 2560
n_channels <- 16
sr <- 256

# Simulate 10 seconds of data with injected artifacts
raw_data <- matrix(rnorm(n_time * n_channels), nrow = n_time, ncol = n_channels)

# Inject a simulated eye-blink artifact into frontal channels
blink <- dnorm(seq(-3, 3, length.out = 100), mean = 0, sd = 1) * 50
blink_onset <- 500
raw_data[blink_onset:(blink_onset + 99), 1:2] <-
  raw_data[blink_onset:(blink_onset + 99), 1:2] + blink

pe <- PhysioExperiment(
  assays = list(raw = raw_data),
  colData = S4Vectors::DataFrame(
    label = paste0("Ch", seq_len(n_channels)),
    type = rep("EEG", n_channels)
  ),
  samplingRate = sr
)
pe
```

## Workflow 1: Built-In ICA

The built-in ICA implementation uses the FastICA algorithm with PCA-based
whitening. It requires no additional packages.

### Step 1: Decompose into Independent Components

`icaDecompose()` separates the multi-channel signal into statistically
independent components. Each component represents a source signal that is
maximally non-Gaussian.

```{r ica-decompose}
ica_result <- icaDecompose(pe, n_components = 16, method = "fastica")
```

The result is a list containing:

- `components`: The independent component time courses (time x components).
- `mixing`: The mixing matrix mapping components to channels.
- `unmixing`: The unmixing matrix mapping channels to components.
- `object`: The updated `PhysioExperiment` with ICA components stored in the
  `"ica_components"` assay.

```{r inspect-result}
str(ica_result$mixing)
str(ica_result$unmixing)

# The PhysioExperiment now contains the ICA components as an assay
pe_ica <- ica_result$object
SummarizedExperiment::assayNames(pe_ica)
```

### Step 2: Identify Artifact Components

Identifying which components represent artifacts is typically done through
visual inspection of component time courses and topographies. Components
corresponding to eye blinks tend to show large, brief deflections concentrated
over frontal electrodes.

```{r identify-artifacts}
# Examine the mixing matrix to see each component's spatial pattern
# Columns of the mixing matrix show the contribution of each component
# to each channel
mixing <- ica_result$mixing
print(round(mixing[, 1:4], 3))

# A component heavily weighted on frontal channels (Ch1, Ch2) with
# characteristic blink morphology is likely an eye-blink artifact
```

### Step 3: Remove Artifact Components

Once artifact components have been identified, `icaRemove()` zeros out those
components and reconstructs the signal from the remaining components.

```{r ica-remove}
# Remove component 1 (identified as eye-blink artifact)
pe_clean <- icaRemove(pe_ica, components = c(1), output_assay = "ica_cleaned")

# The cleaned data is stored in the "ica_cleaned" assay
SummarizedExperiment::assayNames(pe_clean)
```

## Workflow 2: fastICA Package

If the `fastICA` package is installed, you can use the `runICA()` and
`removeICAComponents()` wrapper functions. These provide a simpler interface
that delegates computation to the `fastICA` package.

### Step 1: Run ICA

```{r run-ica}
ica_result2 <- runICA(pe, n_components = 10)
```

The result is a list with:

- `S`: Source matrix (time x components).
- `A`: Mixing matrix (channels x components).
- `W`: Unmixing matrix (components x channels).

```{r inspect-ica}
dim(ica_result2$S)
dim(ica_result2$A)
dim(ica_result2$W)
```

### Step 2: Remove Components

After identifying artifact components, use `removeICAComponents()` to
reconstruct the cleaned signal.

```{r remove-components}
pe_clean2 <- removeICAComponents(pe, ica_result2, remove_components = c(1, 3))
SummarizedExperiment::assayNames(pe_clean2)
```

## Detecting Bad Channels

Before running ICA, it is often helpful to identify and handle bad channels.
Bad channels can degrade ICA decomposition quality because they introduce noise
that may spread across multiple components.

```{r detect-bad}
# Detect channels with abnormally high or low variance
bad_zscore <- detectBadChannels(pe, method = "zscore", threshold = 3)

# Detect channels with low correlation to neighbors
bad_corr <- detectBadChannels(pe, method = "correlation", threshold = 0.4)

# Detect flatline channels
bad_flat <- detectBadChannels(pe, method = "flatline", threshold = 0.01)

# Combine results
bad_all <- unique(c(bad_zscore, bad_corr, bad_flat))
print(bad_all)
```

### Interpolating Bad Channels

Bad channels can be replaced with interpolated values from neighboring
channels before ICA decomposition.

```{r interpolate-bad}
if (length(bad_all) > 0) {
  pe_interp <- interpolateBadChannels(pe, bad_channels = bad_all,
                                       method = "average")
}
```

## Recommended Preprocessing Order

A typical preprocessing pipeline for EEG data incorporating ICA artifact
removal follows this order:

1. **Highpass filter** to remove slow drifts (e.g., 1 Hz cutoff).
2. **Notch filter** to remove power line noise.
3. **Detect and interpolate bad channels**.
4. **Run ICA** on the cleaned, filtered data.
5. **Identify and remove artifact components**.
6. **Re-reference** to average reference.
7. **Epoch** the data around events of interest.
8. **Baseline correct** each epoch.

```{r full-pipeline}
# Step 1: Highpass filter at 1 Hz
pe_hp <- butterworthFilter(pe, low = 1, type = "high")

# Step 2: Notch filter at 50 Hz
pe_notch <- notchFilter(pe_hp, freq = 50)

# Step 3: Detect and interpolate bad channels
bad <- detectBadChannels(pe_notch, method = "zscore")
if (length(bad) > 0) {
  pe_notch <- interpolateBadChannels(pe_notch, bad_channels = bad)
}

# Step 4: Run ICA
ica <- icaDecompose(pe_notch, method = "fastica")
pe_ica <- ica$object

# Step 5: Remove artifact components (after visual inspection)
pe_clean <- icaRemove(pe_ica, components = c(1))

# Step 6: Re-reference to average
pe_ref <- rereference(pe_clean, ref_type = "average")
```

## Choosing the Number of Components

The number of ICA components to extract affects decomposition quality:

- **Too few components**: Artifact and neural sources may be mixed within the
  same component, making clean separation impossible.
- **Too many components**: Components may capture noise rather than meaningful
  sources, and the decomposition may become unstable.

A common practice is to use the same number of components as channels, or to
reduce dimensionality first via PCA (which `icaDecompose()` does internally
during the whitening step).

```{r n-components}
# Use all channels
ica_full <- icaDecompose(pe, n_components = n_channels)

# Use a reduced number (e.g., for rank-deficient data after average reference)
ica_reduced <- icaDecompose(pe, n_components = 12)
```

## Summary

This vignette demonstrated two ICA workflows available in PhysioPreprocess:

- **Built-in ICA** via `icaDecompose()` and `icaRemove()`, requiring no
  additional packages.
- **fastICA-based** via `runICA()` and `removeICAComponents()`, wrapping the
  `fastICA` package.

Both workflows follow the same general pattern: decompose the signal,
identify artifact components, and reconstruct the cleaned signal with those
components removed.

For general preprocessing operations (filtering, resampling, re-referencing,
detrending), see `vignette("preprocessing-guide", package = "PhysioPreprocess")`.

## References

- Hyvarinen A, Oja E (2000). "Independent component analysis: algorithms and
  applications." *Neural Networks*, 13(4-5), 411-430.
- Makeig S, Bell AJ, Jung TP, Sejnowski TJ (1996). "Independent component
  analysis of electroencephalographic data." *Advances in Neural Information
  Processing Systems*, 8, 145-151.
- Delorme A, Makeig S (2004). "EEGLAB: an open source toolbox for analysis of
  single-trial EEG dynamics including independent component analysis."
  *Journal of Neuroscience Methods*, 134(1), 9-21.
